---
format: gfm
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.path: "man/figures/README-"
    out.width: "100%"
    collapse: true
    comment: "#>"
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# itscausal

<!-- badges: start -->
<!-- badges: end -->

`itscausal` estimates the causal effect of a policy intervention on a panel of
units observed over time — an **interrupted time series (ITS)** design.

The key idea: train ML models on the **pre-intervention** window to predict each
unit's counterfactual outcome (what would have happened without treatment). The
individual treatment effect (ITE) is the difference between observed and
counterfactual outcomes:

$$\text{ITE}_{it} = y_{it} - \hat{y}_{it}^{(0)}, \quad t > 0$$

## Installation

```r
# install.packages("devtools")
devtools::install_github("ASallin/itscausal")
```

## Workflow

```{r setup, message=FALSE, warning=FALSE}
devtools::load_all(quiet = TRUE)
library(data.table)
library(ggplot2)

set.seed(20260101)
```

### 1. Simulate a panel

60 individuals × 84 months; intervention at month 60; 24 post-intervention periods; true level-shift = **−3**.

```{r simulate}
n_id        <- 60
n_time      <- 84
INDEX       <- 60L
true_effect <- -3

ar1 <- function(n, phi = 0.6, sigma = 1) {
  e <- rnorm(n, 0, sigma); x <- numeric(n); x[1] <- e[1]
  for (t in 2:n) x[t] <- phi * x[t - 1] + e[t]; x
}

X     <- sample(c(0, 1), n_id, replace = TRUE)
id_fx <- rnorm(n_id, 0, 3)

df <- data.table(
  ID    = rep(seq_len(n_id), each = n_time),
  time  = rep(seq_len(n_time), n_id),
  X     = rep(X, each = n_time),
  id_fx = rep(id_fx, each = n_time)
)

season_coefs <- c(0, 1.5, 2.8, 3.1, 2.4, 1.2, -0.5, -1.8, -2.1, -1.4, 0, 0.3)
df[, month  := ((time - 1) %% 12) + 1]
df[, year   := ((time - 1) %/% 12) + 1]
df[, season := season_coefs[month]]
df[, error  := unlist(replicate(n_id, ar1(n_time, phi = 0.5, sigma = 1.5),
                                simplify = FALSE))]
df[, post   := as.integer(time >= INDEX)]
df[, y      := 2 + 0.05 * time + 0.8 * X + season + id_fx +
               true_effect * post + error]
```

```{r fig-rawdata, echo=FALSE}
avg <- df[, .(mean_y = mean(y), sd_y = sd(y)), by = time]
avg[, time_c := time - INDEX]

ggplot(avg, aes(x = time_c, y = mean_y)) +
  geom_ribbon(aes(ymin = mean_y - sd_y, ymax = mean_y + sd_y),
              alpha = 0.15, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "firebrick") +
  annotate("text", x = 1, y = max(avg$mean_y) - 0.5,
           label = "Intervention", hjust = 0, colour = "firebrick", size = 3.5) +
  labs(x = "Time relative to intervention (months)", y = "Mean outcome (\u00b1 1 SD)") +
  theme_classic()
```

### 2. Forecast the counterfactual — comparing models

`forecastITS()` trains an ensemble of ML models on each cross-validation fold
and forecasts the post-intervention counterfactual.
Here we compare three specifications: linear model only, XGBoost only, and the
ensemble of both.

```{r forecast, message=FALSE, warning=FALSE}
n_post <- n_time - INDEX   # 24 post-intervention periods

fore_lm <- forecastITS(
  data            = copy(df),
  time            = "time",
  key             = "ID",
  y               = "y",
  INDEX           = INDEX,
  WINDOW          = 12L,
  STEPS           = 6L,
  FORECASTUNITS   = n_post,
  covariates_time = c("month", "year"),
  covariates_fix  = "X",
  method          = c("lm"),
  K               = 5L
)

fore_xgb <- forecastITS(
  data            = copy(df),
  time            = "time",
  key             = "ID",
  y               = "y",
  INDEX           = INDEX,
  WINDOW          = 12L,
  STEPS           = 6L,
  FORECASTUNITS   = n_post,
  covariates_time = c("month", "year"),
  covariates_fix  = "X",
  method          = c("xgboost"),
  K               = 5L
)

fore_ens <- forecastITS(
  data            = copy(df),
  time            = "time",
  key             = "ID",
  y               = "y",
  INDEX           = INDEX,
  WINDOW          = 12L,
  STEPS           = 6L,
  FORECASTUNITS   = n_post,
  covariates_time = c("month", "year"),
  covariates_fix  = "X",
  method          = c("lm", "xgboost", "rf"),
  K               = 5L
)

# Ensemble RMSE weights
round(fore_ens$RMSEweights, 3)
```

```{r fig-forecast, echo=FALSE}
tidy_out <- function(fore, label) {
  d <- data.table(fore$out)
  setnames(d, c("ID", "time_c", "y_hat"))
  d[, model := label]
  d
}

preds <- rbind(
  tidy_out(fore_lm,  "lm"),
  tidy_out(fore_xgb, "xgboost"),
  tidy_out(fore_ens, "ensemble")
)

obs <- df[, .(y = mean(y)), by = .(time_c = time - INDEX)]
preds_avg <- preds[, .(y_hat = mean(y_hat, na.rm = TRUE)), by = .(time_c, model)]
preds_avg <- na.omit(preds_avg)   # drop periods where all folds returned NA

ggplot(obs, aes(x = time_c, y = y)) +
  geom_line(colour = "black", linewidth = 0.7) +
  geom_line(data = preds_avg[time_c > 0],
            aes(y = y_hat, colour = model, linetype = model), linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  scale_colour_manual(values = c(lm = "#1b7837", xgboost = "#762a83", ensemble = "#d6604d")) +
  labs(x = "Time relative to intervention", y = "Mean outcome",
       colour = "Model", linetype = "Model",
       caption = "Black = observed; coloured = counterfactual forecast by model") +
  theme_classic() +
  theme(legend.position = "bottom")
```

### 3. Individual & average treatment effects

```{r ate, message=FALSE, warning=FALSE}
ite_lm  <- iteITS(fore_lm)
ite_xgb <- iteITS(fore_xgb)
ite_ens <- iteITS(fore_ens)

n_post <- max(ite_ens$ites$time)

summarise_tate <- function(fore, ite, label) {
  ate <- ateITS(fore, ite, n.periods = n_post)
  data.frame(
    Model = label,
    TATE = round(ate$TATE$pred$ite, 3),
    SD = round(ate$TATE$sd$sd, 3)
  )
}

results <- rbind(
  summarise_tate(fore_lm,  ite_lm,  "lm"),
  summarise_tate(fore_xgb, ite_xgb, "xgboost"),
  summarise_tate(fore_ens, ite_ens, "ensemble")
)
results$lower <- round(results$TATE - 1.96 * results$SD, 3)
results$upper <- round(results$TATE + 1.96 * results$SD, 3)
results
```

```{r fig-ite, echo=FALSE}
ite_pop_ens <- as.data.frame(ite_ens$ites)[, c("time", "ite", "sd")]
ite_pop_ens <- aggregate(cbind(ite, sd) ~ time, data = ite_pop_ens, FUN = mean)
mean_est    <- mean(ite_pop_ens$ite)

ggplot(ite_pop_ens, aes(x = time, y = ite)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_errorbar(aes(ymin = ite - sd, ymax = ite + sd), width = 0.3) +
  geom_hline(yintercept = true_effect, linetype = "dashed",
             colour = "firebrick", linewidth = 0.8) +
  geom_hline(yintercept = mean_est, linetype = "dashed",
             colour = "grey40", linewidth = 0.8) +
  annotate("text", x = max(ite_pop_ens$time) - 0.3, y = true_effect - 0.3,
           label = paste("True =", true_effect),
           colour = "firebrick", hjust = 1, size = 3.5) +
  annotate("text", x = max(ite_pop_ens$time) - 0.3, y = mean_est + 0.3,
           label = paste("Mean est. =", round(mean_est, 2)),
           colour = "grey40", hjust = 1, size = 3.5) +
  labs(x = "Post-intervention period",
       y = "ITE (observed \u2212 counterfactual)",
       caption = "Ensemble model; bars = mean ITE per period \u00b1 1 SD") +
  theme_classic()
```
