---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)


# Generate AR Process
# Generate the time series
ar_process <- function(n_time, ar_params) {
  errors <- rnorm(n_time, 0, 1) # Generate the random errors
  simulated_data <- numeric(n_time)

  for (i in length(ar_params):n_time) {
    simulated_data[i] <- sum(ar_params * simulated_data[(i - length(ar_params)):(i - 1)]) + errors[i]
  }

  simulated_data
}

ar_params <- c(
  0.5, 0.3, 0.001, 0.2,
  0, -.050, -.020, 0.001, -0.02,
  0, 0.03, 0.15
) # Autocorrelation parameters with order p
order <- 12 # Order of the autocorrelation


library(lintr)

# Set linter options
linters <- linters_with_defaults(
  line_length_linter = line_length_linter(120),
  object_name_linter = object_name_linter(styles = "camelCase"),
  spaces_left_parentheses_linter = NULL
)
# lint_package(path = ".")
```

# itscausal

<!-- badges: start -->
<!-- badges: end -->

Welcome to the package page of itscausal. The goal of itscausal is to flexibly estimate
interrupted time series With causal machine learning.

## Installation

You can install the development version of itscausal from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ASallin/itscausal", force = TRUE)
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r simulation, echo=TRUE, message=FALSE, warning=FALSE}
library(itscausal)
library(data.table)
library(dplyr)
library(ggplot2)

set.seed(23493)


ar_process <- function(n, phi = 0.7, sigma = 1) {
  e <- rnorm(n, 0, sigma)
  x <- numeric(n)
  x[1] <- e[1]
  for (t in 2:n) {
    x[t] <- phi * x[t - 1] + e[t]
  }
  return(x)
}

arma_process <- function(n, ar = 0.7, ma = 0.3, sigma = 1) {
  e <- rnorm(n, 0, sigma)
  x <- numeric(n)
  x[1] <- e[1]
  for (t in 2:n) {
    x[t] <- ar * x[t - 1] + e[t] + ma * e[t - 1]
  }
  return(x)
}

heteroskedastic_errors <- function(n, base_sigma = 1, increase_rate = 0.1) {
  sigma <- base_sigma + increase_rate * (1:n)
  return(rnorm(n, 0, sigma))
}

set.seed(23493)

# Generate simulated dataset
n_time <- 100 # Number of time points
intervention <- round(0.8 * n_time) # Time point of intervention
n_id <- 60 # Number of unique individuals
constant <- 1

X <- sample(c(0, 1), n_id, T)
param_timetrend <- 0.07
param_dummyintervention <- -0.15
param_slopeintervention <- -0.10

# Seasonality
season_effect <- c(
  0,
  rnorm(2, 0, 4),
  rnorm(4, -2, 0.5),
  rnorm(3, 0, 1)
)

# Simulate an ITS
df <- data.frame(
  ID = rep(seq(1:n_id), each = n_time),
  X = rep(X, each = n_time),
  season = rep(
    c(rep(1:12, n_time %/% 12), (1:12)[1:(n_time %% 12)]),
    n_id
  ),
  id.effect = rep(rnorm(n_id, 0, 10), each = n_time),
  season_effect = rep(season_effect, times = n_time),
  post = rep(c(rep(0, intervention), rep(1, 0.2 * n_time)), n_id),
  time = rep((1:n_time) - intervention, n_id),
  error_simple = rnorm(n_time * n_id, 0, 2),
  error_ar = unlist(replicate(n_id, ar_process(n_time, sigma = 2), simplify = FALSE)),
  error_arma = unlist(replicate(n_id, arma_process(n_time), simplify = FALSE)),
  error_heteroskedastic = unlist(replicate(n_id, heteroskedastic_errors(n_time), simplify = FALSE))
)

df$abs_time <- df$time + abs(min(df$time))

# Generate y with different error structures
df$y_simple <- with(
  df,
  constant + param_timetrend * abs_time
    + (param_dummyintervention * post)
    + param_slopeintervention * post * abs_time
    + 0.2 * X
    + season_effect
    + error_simple
)

df$y_ar <- with(
  df,
  constant + param_timetrend * abs_time
    + (param_dummyintervention * post)
    + param_slopeintervention * post * abs_time
    + 0.2 * X
    + season_effect
    + error_ar
)

df$y_arma <- with(
  df,
  constant + param_timetrend * abs_time
    + (param_dummyintervention * post)
    + param_slopeintervention * post * abs_time
    + 0.2 * X
    + season_effect
    + error_arma
)

df$y_heteroskedastic <- with(
  df,
  constant + param_timetrend * abs_time
    + (param_dummyintervention * post)
    + param_slopeintervention * post * abs_time
    + 0.2 * X
    + season_effect
    + error_heteroskedastic
)

df <- data.table(df)


# Simulate y with seasonal effects
df <- df[
  , `:=`(
    model = constant
    + param_timetrend * (abs_time)
      + season_effect
      + 0.2 * X,
    forecast = constant
    + param_timetrend * abs_time
      + param_dummyintervention * post
      + param_slopeintervention * abs_time * post
      + season_effect
      + 0.2 * X
      + 0.03 * X * param_slopeintervention
  )
]
```

```{r graph, echo = FALSE}
# Graph
df %>%
  group_by(time, post) %>%
  summarise(
    y_simple = mean(y_simple),
    y_ar = mean(y_ar),
    y_arma = mean(y_arma),
    y_heteroskedastic = mean(y_heteroskedastic),
    model = mean(model),
    forecast = mean(forecast)
  ) %>%
  # mutate(y = ifelse(post == 1, NA, y)) |>
  ggplot(aes(y = y_simple, x = time)) +
  geom_line(col = "#1c1cdd") +
  geom_line(aes(y = model), color = "#fa1420") +
  geom_line(aes(y = forecast), color = "#066613", size = 1) +
  geom_line(aes(y = y_ar), color = "#030366") +
  geom_line(aes(y = y_arma), color = "#ad5a5e66") +
  geom_line(aes(y = y_heteroskedastic), color = "#bb49be66") +
  labs(
    x = "Time to/from the intervention",
    y = "Average y per time period"
  ) +
  geom_vline(xintercept = 0, color = "red", linetype = 2) +
  theme_classic()
scale_x_continuous(lim = c(-36, 24))
```


We can visualize the its as follows:

```{r}
# Compute effects
df <- df[, ite := ifelse(post == 0, NA, model - forecast)]
ate5 <- mean(df[time < 6 & post == 1, ]$ite)
ate1 <- mean(df[time < 2 & post == 1, ]$ite)

print(ate5)
print(ate1)
```


We can reproduce these results using the `itscausal` package.

```{r itscausal}
window <- 36L

fore_y_simple <- forecastITS(
  data = df,
  time = "time",
  INDEX = 0L,
  WINDOW = window,
  STEPS = 5,
  covariates_time = c("season"),
  covariates_fix = c("X"),
  key = "ID",
  y = "y_simple",
  method = c("lm", "xgboost"),
  K = 5
)

fore_y_ar <- forecastITS(
  data = df,
  time = "time",
  INDEX = 0L,
  WINDOW = window,
  STEPS = 5,
  covariates_time = c("season"),
  covariates_fix = c("X"),
  key = "ID",
  y = "y_ar",
  method = c("lm", "xgboost"),
  K = 5
)

fore_y_arma <- forecastITS(
  data = df,
  time = "time",
  INDEX = 0L,
  WINDOW = window,
  STEPS = 5,
  covariates_time = c("season"),
  covariates_fix = c("X"),
  key = "ID",
  y = "y_arma",
  method = c("lm", "xgboost"),
  K = 5
)

fore_y_heteroskedastic <- forecastITS(
  data = df,
  time = "time",
  INDEX = 0L,
  WINDOW = window,
  STEPS = 5,
  covariates_time = c("season"),
  covariates_fix = c("X"),
  key = "ID",
  y = "y_heteroskedastic",
  method = c("lm", "xgboost"),
  K = 5
)

dfFinal_simple <- fore_y_simple$out
dfFinal_ar <- fore_y_ar$out
dfFinal_arma <- fore_y_arma$out
dfFinal_heteroskedastic <- fore_y_heteroskedastic$out
```


```{r graphML, echo = FALSE}
dfFinal_simple %>%
  right_join(df) %>%
  group_by(time) %>%
  summarise(y = mean(y_simple), y.pred = mean(y_simple_hat)) %>%
  ggplot(aes(y = y, x = time)) +
  geom_line() +
  geom_line(aes(y = y.pred, x = time), color = "red") +
  geom_line(
    aes(
      x = time,
      y = constant + param_timetrend * (time + abs(min(df$time)))
        + season_effect
    ),
    col = "#e95a08", linetype = 3, size = 1
  ) +
  labs(
    x = "Time to/from the intervention",
    y = "Average y per time period"
  ) +
  geom_vline(xintercept = 0, color = "red", linetype = 2) +
  theme_classic()


dfFinal_ar %>%
  right_join(df) %>%
  group_by(time) %>%
  summarise(y = mean(y_ar), y.pred = mean(y_ar_hat)) %>%
  ggplot(aes(y = y, x = time)) +
  geom_line() +
  geom_line(aes(y = y.pred, x = time), color = "red") +
  geom_line(
    aes(
      x = time,
      y = constant + param_timetrend * (time + abs(min(df$time)))
        + season_effect
    ),
    col = "#e95a08", linetype = 3, size = 1
  ) +
  labs(
    x = "Time to/from the intervention",
    y = "Average y per time period"
  ) +
  geom_vline(xintercept = 0, color = "red", linetype = 2) +
  theme_classic()


dfFinal_arma %>%
  right_join(df) %>%
  group_by(time) %>%
  summarise(y = mean(y_arma), y.pred = mean(y_arma_hat)) %>%
  ggplot(aes(y = y, x = time)) +
  geom_line() +
  geom_line(aes(y = y.pred, x = time), color = "red") +
  geom_line(
    aes(
      x = time,
      y = constant + param_timetrend * (time + abs(min(df$time)))
        + season_effect
    ),
    col = "#e95a08", linetype = 3, size = 1
  ) +
  labs(
    x = "Time to/from the intervention",
    y = "Average y per time period"
  ) +
  geom_vline(xintercept = 0, color = "red", linetype = 2) +
  theme_classic()

dfFinal_heteroskedastic %>%
  right_join(df) %>%
  group_by(time) %>%
  summarise(y = mean(y_heteroskedastic), y.pred = mean(y_heteroskedastic_hat)) %>%
  ggplot(aes(y = y, x = time)) +
  geom_line() +
  geom_line(aes(y = y.pred, x = time), color = "red") +
  geom_line(
    aes(
      x = time,
      y = constant + param_timetrend * (time + abs(min(df$time)))
        + season_effect
    ),
    col = "#e95a08", linetype = 3, size = 1
  ) +
  labs(
    x = "Time to/from the intervention",
    y = "Average y per time period"
  ) +
  geom_vline(xintercept = 0, color = "red", linetype = 2) +
  theme_classic()
```

```{r ite}
# Now we need a way to compute the ITE
iteM <- iteITS(forecast.object = fore)
```


## Compute the instantaneous treatment effect for the whole population
```{r InstATE}
InstAte <- ateITS(fore, ite.object = iteM)
InstATE <- InstAte$InstATE

ggplot(InstATE$pred, aes(x = time, y = ite)) +
  geom_bar(stat = "identity")
```


## Per groupdf
```{r ate}
ate1its <- ateITS(fore, iteM, n.periods = 1)
paste("mean = ", round(ate1its$TATE$pred$ite, 3), "; sd = ", round(ate1its$TATE$sd$sd, 3))

ate5its <- ateITS(fore, iteM, n.periods = 5)
paste("mean = ", round(ate5its$TATE$pred$ite, 3), "; sd = ", round(ate5its$TATE$sd$sd, 3))

ate1
ate5
```