---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# itscausal

<!-- badges: start -->
<!-- badges: end -->

`itscausal` estimates the causal effect of a policy intervention on a panel of
units observed over time — an **interrupted time series (ITS)** design.

The key idea: train ML models on the **pre-intervention** window to predict each
unit's counterfactual outcome (what would have happened without treatment). The
individual treatment effect (ITE) is the difference between observed and
counterfactual outcomes.

## Installation

```r
# install.packages("devtools")
devtools::install_github("ASallin/itscausal")
```

## Workflow

```{r setup, message=FALSE, warning=FALSE}
library(itscausal)
library(data.table)
library(ggplot2)

set.seed(20240101)
```

### 1. Simulate a panel

60 individuals × 72 months; intervention at month 60; true level-shift effect = **−3**.

```{r simulate}
n_id        <- 60
n_time      <- 72
INDEX       <- 60L
true_effect <- -3

ar1 <- function(n, phi = 0.6, sigma = 1) {
  e <- rnorm(n, 0, sigma); x <- numeric(n); x[1] <- e[1]
  for (t in 2:n) x[t] <- phi * x[t - 1] + e[t]; x
}

X     <- sample(c(0, 1), n_id, replace = TRUE)
id_fx <- rnorm(n_id, 0, 3)

df <- data.table(
  ID    = rep(seq_len(n_id), each = n_time),
  time  = rep(seq_len(n_time), n_id),
  X     = rep(X, each = n_time),
  id_fx = rep(id_fx, each = n_time)
)

season_coefs <- c(0, 1.5, 2.8, 3.1, 2.4, 1.2, -0.5, -1.8, -2.1, -1.4, 0, 0.3)
df[, month  := ((time - 1) %% 12) + 1]
df[, season := season_coefs[month]]
df[, error  := unlist(replicate(n_id, ar1(n_time, phi = 0.5, sigma = 1.5),
                                simplify = FALSE))]
df[, post   := as.integer(time >= INDEX)]
df[, y      := 2 + 0.05 * time + 0.8 * X + season + id_fx +
               true_effect * post + error]
df[, time_c := time - INDEX]
```

### 2. Forecast the counterfactual

```{r forecast, message=FALSE, warning=FALSE}
fore <- forecastITS(
  data            = df,
  time            = "time_c",
  INDEX           = 0L,
  WINDOW          = 12L,
  STEPS           = 3,
  covariates_time = "season",
  covariates_fix  = "X",
  key             = "ID",
  y               = "y",
  method          = c("lm", "xgboost"),
  K               = 5
)
```

```{r fig-forecast, echo=FALSE}
pred_df <- do.call(rbind, lapply(fore$forecast, function(x) x$y_hat))
pred_long <- data.table::melt(data.table(pred_df), id.vars = c("ID", "time"),
                               variable.name = "step", value.name = "y_hat")
pred_long <- pred_long[!is.na(y_hat)]
pred_agg  <- pred_long[, .(y_hat = mean(y_hat)), by = time]

obs_agg <- df[, .(y = mean(y)), by = .(time = time_c)]

ggplot(obs_agg, aes(x = time, y = y)) +
  geom_line(colour = "steelblue") +
  geom_line(data = pred_agg[time > 0], aes(y = y_hat),
            colour = "firebrick", linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(x = "Time relative to intervention",
       y = "Mean outcome",
       caption = "Blue = observed; red dashed = counterfactual forecast") +
  theme_classic()
```

### 3. Individual treatment effects

```{r ite}
iteM <- iteITS(forecast.object = fore)
```

### 4. Average treatment effects

```{r ate}
n_post    <- max(iteM$ites$time)
ate_total <- ateITS(fore, iteM, n.periods = n_post)

cat(sprintf(
  "TATE over %d periods:  %.3f  (SD = %.3f)\nTrue effect:           %.1f\n",
  n_post,
  ate_total$TATE$pred$ite,
  ate_total$TATE$sd$sd,
  true_effect
))
```

```{r fig-ite, echo=FALSE}
ite_pop  <- as.data.frame(iteM$ites)[, c("time", "ite", "sd")]
ite_pop  <- aggregate(cbind(ite, sd) ~ time, data = ite_pop, FUN = mean)
mean_est <- mean(ite_pop$ite)

ggplot(ite_pop, aes(x = time, y = ite)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_errorbar(aes(ymin = ite - sd, ymax = ite + sd), width = 0.3) +
  geom_hline(yintercept = true_effect, linetype = "dashed",
             colour = "firebrick", linewidth = 0.8) +
  geom_hline(yintercept = mean_est, linetype = "dashed",
             colour = "grey40", linewidth = 0.8) +
  annotate("text", x = max(ite_pop$time) - 0.3, y = true_effect - 0.25,
           label = paste("True =", true_effect),
           colour = "firebrick", hjust = 1, size = 3.5) +
  annotate("text", x = max(ite_pop$time) - 0.3, y = mean_est + 0.25,
           label = paste("Mean est. =", round(mean_est, 2)),
           colour = "grey40", hjust = 1, size = 3.5) +
  labs(x = "Post-intervention period",
       y = "ITE (observed \u2212 counterfactual)") +
  theme_classic()
```
